# Laboratory performance

## Load data

```{r load_counts, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
sample_metadata <- read_tsv(paste0("data/metadata.tsv")) %>%
    mutate(Extraction=factor(Extraction, levels=c("ZYMO","DREX","EHEX"))) %>%
    rename(dataset=Dataset)
```

```{r extraction_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
extraction_data <- airtable("tblBcTZcRG1E9wsGO", "appQpr6MxnaiVHsHy") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("EX DNA ng","Datasets_flat"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  rename(id=1,extract=2,dataset=3) %>%
  filter(dataset %in% sample_metadata$dataset) %>%
  select(dataset,extract)
```

```{r library_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
library_data <- airtable("tblo6AuYpxbbGw9gh", "appQpr6MxnaiVHsHy") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("Required PCR cycles","Datasets_flat"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  rename(id=1,pcr=2,dataset=3) %>%
  filter(dataset %in% sample_metadata$dataset) %>%
  select(dataset,pcr)
```

```{r indexing_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
indexing_data <- airtable("tblhfsiR4NI9XJQG0", "appQpr6MxnaiVHsHy") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("Adaptors (nM)","Library (nM)","Datasets_flat"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  rename(id=1,adaptors=2,library=3,dataset=4) %>%
  filter(dataset %in% sample_metadata$dataset) %>%
  mutate(ratio=library/(adaptors+library)) %>%
  select(dataset,adaptors,library,ratio)
```

```{r preprocessing_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
preprocessing_data <- airtable("tblJfLRU2FIVz37Y1", "appQpr6MxnaiVHsHy") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("bases_pre_fastp","bases_post_fastp","host_bases","metagenomic_bases","singlem_fraction","C","EHI_plaintext"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  rename(id=1,nonpareil=C,dataset=EHI_plaintext,microbial_fraction=singlem_fraction) %>%
  filter(dataset %in% sample_metadata$dataset) %>%
  select(dataset,bases_pre_fastp,bases_post_fastp,bases_post_fastp,metagenomic_bases,microbial_fraction,nonpareil)
```

```{r assembly_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
assembly_data <- airtable("tblG6ZIvkYN844I97", "appQpr6MxnaiVHsHy") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("EHI_number_api","AB_batch_api","assembly_length","N50","L50","num_contigs","num_bins","metagenomic_bases"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  filter(AB_batch_api %in% c("ABB0409","ABB0410","ABB0411","ABB0412","ABB0413","ABB0414","ABB0415","ABB0416","ABB0417","ABB0418","ABB0419","ABB0420")) %>% 
  select(-AB_batch_api) %>% 
  rename(id=1,assembly_mapped_bases=metagenomic_bases,dataset=EHI_number_api) %>%
  filter(dataset %in% sample_metadata$dataset) %>%
  select(dataset,assembly_length,N50,L50,num_contigs,num_bins,assembly_mapped_bases)
```

```{r mapping_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
mapping_data <- airtable("tblWDyQmM9rQ9wq57", "appWbHBNLE6iAsMRV") %>% #get base ID from Airtable browser URL
  read_airtable(., fields = c("DM_batch_static","EHI_sample_static","MAG_mapping_percentage"), id_to_col = TRUE) %>% #get 3 columns from MAGs table
  filter(DM_batch_static %in% c("DMB0121","DMB0122","DMB0123","DMB0124","DMB0125","DMB0126","DMB0127","DMB0128","DMB0129","DMB0130","DMB0131","DMB0132","DMB0133","DMB0134")) %>% 
  mutate(catalogue = case_when(
    DM_batch_static == "DMB0134" ~ "All",
    DM_batch_static %in% c("DMB0121","DMB0122","DMB0123","DMB0124","DMB0125",
                "DMB0126","DMB0127","DMB0128","DMB0129","DMB0130",
                "DMB0131","DMB0132") ~ "Species",
    TRUE ~ NA_character_
  )) %>% 
  rename(id=1,dataset=EHI_sample_static) %>%
  filter(dataset %in% sample_metadata$dataset) %>%
  select(dataset,MAG_mapping_percentage,catalogue)
```

```{r genomics_data, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
genomics_data <- read_tsv("data/host_genomics_stats.tsv")
```

```{r data_merge, warning=FALSE, comments="", message=FALSE, echo=FALSE, eval=FALSE}
all_data <- sample_metadata %>%
    left_join(extraction_data, by=join_by(dataset==dataset)) %>%
    left_join(library_data, by=join_by(dataset==dataset)) %>%
    left_join(indexing_data, by=join_by(dataset==dataset)) %>%
    left_join(preprocessing_data, by=join_by(dataset==dataset)) %>%
    left_join(assembly_data, by=join_by(dataset==dataset)) %>%
    left_join(mapping_data, by=join_by(dataset==dataset)) %>%
    left_join(genomics_data, by=join_by(dataset==dataset))

write_tsv(all_data, file="data/all_data.tsv")
```

Load data previously compiled from the database:

```{r load_laboroty_data, warning=FALSE, comments="", message=FALSE}
all_data <- read_tsv("data/all_data.tsv") %>%
    mutate(Taxon=factor(Taxon,levels=c("Amphibian",
                                       "Reptile",
                                       "Mammal",
                                       "Bird",
                                       "Control"))) %>%
    mutate(Extraction=factor(Extraction,levels=c("ZYMO",
                                                 "DREX",
                                                 "EHEX"))) %>%
    mutate(Species=factor(Species,levels=c("Calotriton asper",
                                           "Lissotriton helveticus",
                                           "Salamandra atra",
                                           "Chalcides striatus",
                                           "Natrix astreptophora",
                                           "Podarcis muralis",
                                           "Plecotus auritus",
                                           "Sciurus carolinensis",
                                           "Trichosurus vulpecula",
                                           "Geospizopsis unicolor",
                                           "Perisoreus infaustus",
                                           "Zonotrichia capensis",
                                           "Extraction control",
                                           "Library control")))
```

## DNA yield

Total amount of DNA extracted from the 150 ul subset of the bead-beaten sample. 

```{r yield_summary, warning=FALSE, comments="", message=FALSE}
all_data %>%
    select(Extraction,extract,Taxon) %>%
    group_by(Taxon,Extraction) %>%
    summarise(value = sprintf("%.0f±%.0f", mean(extract), sd(extract))) %>%
    pivot_wider(names_from = Extraction, values_from = value) %>%
    tt(caption = "Mean and standard deviation of total DNA nanograms")
```

```{r yield_plot, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
all_data %>%
    select(Library,Species,Extraction,extract,Taxon) %>%
    unique() %>%
    ggplot(aes(x=Extraction, y=extract, color=Species, group=Extraction))+ 
        geom_boxplot(outlier.shape = NA, fill="#f4f4f4", color="#8c8c8c") + 
        geom_jitter() + 
        scale_color_manual(values=vertebrate_colors) +
        facet_grid(. ~ Taxon, scales = "free") +
        theme_minimal() +
        labs(y="DNA yield (ng)",x="Extraction method")
```

```{r yield_test, warning=FALSE, comments="", message=FALSE}
all_data  %>%
    filter(Taxon != "Control") %>%
    lmerTest::lmer(extract ~ Extraction + (1 | Sample) + (1 | Species), data = ., REML = FALSE) %>%
    broom.mixed::tidy() %>%
    tt()
```

> DNA yields vary across host taxa.
> - Amphibians and mammals yield the highest total amounts of DNA.
> - Birds yield the lowest amounts of DNA.
> - Controls yield barely quantifiable DNA.
> EHEX yields highest amounts of total DNA.
> - DNA recovery raks of extraction methods are maintained across taxa.

## Library performance

Number of PCR cycles required for reaching the plateau phase of the indexing PCR. A higher number indicates a lower amount of amplifiable library template. When weighed by the input DNA, the required number of PCR cycles serves as an estimator of library performance.

```{r pcr_summary, warning=FALSE, comments="", message=FALSE}
all_data %>%
    group_by(Taxon,Extraction) %>%
    summarise(value = sprintf("%.1f±%.1f", mean(pcr), sd(pcr))) %>%
    pivot_wider(names_from = Extraction, values_from = value) %>%
    tt(caption = "Mean and standard deviation of optimal number of PCR cycles")
```

```{r pcr_plot, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
all_data %>%
    select(Library,Species,Extraction,pcr,Taxon) %>%
    unique() %>%
    ggplot(aes(x=Extraction,y=pcr, color=Species, group=Extraction))+ 
        scale_y_reverse() +
        geom_boxplot(outlier.shape = NA, fill="#f4f4f4", color="#8c8c8c") + 
        geom_jitter() + 
        scale_color_manual(values=vertebrate_colors) +
        facet_grid(. ~ Taxon, scales = "free") +
        theme_minimal() +
        labs(y="Optimal number of PCR cycles",x="Extraction method")
```

```{r pcr_test, warning=FALSE, comments="", message=FALSE}
all_data  %>%
    filter(Taxon != "Control") %>%
    lmerTest::lmer(pcr ~ Extraction + (1 | Sample) + (1 | Species), data = ., REML = FALSE) %>%
    broom.mixed::tidy() %>%
    tt()
```
