# Metagenomic data

## Library complexity

Nonpareil estimate of the metagenomic complexity after removing host DNA.

```{r nonpareil_summary, warning=FALSE, comments="", message=FALSE}
all_data %>%
    select(dataset,Extraction,nonpareil,Taxon) %>%
    unique() %>%
    group_by(Taxon,Extraction) %>%
    summarise(value = sprintf("%.1f±%.1f", mean(nonpareil), sd(nonpareil))) %>%
    pivot_wider(names_from = Extraction, values_from = value) %>%
    tt(caption = "Mean and standard deviation of breadth of host genome coverage")
```

```{r nonpareil_plot, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
all_data %>%
    select(dataset,Extraction,nonpareil,Taxon,Species) %>%
    unique() %>%
    ggplot(aes(x=Extraction,y=nonpareil, color=Species, group=Extraction)) + 
        geom_boxplot(outlier.shape = NA, fill="#f4f4f4", color="#8c8c8c") + 
        geom_jitter() + 
        scale_color_manual(values=vertebrate_colors) +
        facet_grid(. ~ Taxon, scales = "free") +
        theme_minimal() +
        labs(y="Nonpareil completeness",x="Extraction method")
```

```{r nonpareil_test, warning=FALSE, comments="", message=FALSE}
all_data  %>%
    select(dataset,Extraction,Sample,Species,nonpareil,Taxon,catalogue) %>%
    unique() %>%
    filter(Taxon != "Control") %>%
    #filter(catalogue != "All") %>%
    MASS::glmmPQL(nonpareil~ Taxon+Extraction,random=~1|Species/Sample,
               family="quasibinomial",data=.) %>%
    car::Anova(.,test.statistic="Wald")

data<-all_data  %>%
    select(dataset,Extraction,Sample,Species,nonpareil,Taxon,catalogue) %>%
    unique() %>%
    filter(Taxon != "Control") #%>% filter(catalogue != "All")
M<-MASS::glmmPQL(nonpareil~ Taxon+Extraction,random=~1|Species/Sample,
               family="quasibinomial",data=data)

Anova_table<-car::Anova(M,test.statistic="Wald",type="III")%>%
    mutate(F=Chisq/Df) # Approximate F value
n <- nrow(M$data)  # Total number of observations
df_fixed <- length(fixef(M))  # Number of fixed effect parameters
df_random <- length(ranef(M))  # Number of random effect parameters (or levels)
approx_residual_df <- n - df_fixed - df_random # Approx. residual Df
data.frame(Anova_table,
           p_value=pf(Anova_table$F, Anova_table$Df, approx_residual_df, lower.tail = FALSE))%>%
  select(F,p_value)

broom.mixed::tidy(M) %>%
    tt()

plot_model(M,type="pred",terms=c("Taxon","Extraction"))

VarCorr(M)

# Within-taxon between-species variance: 0%
4*(2.092129e-09/(2.092129e-09+1.658436e-01))
# Within-species between-sample variance: 4%
4*(1.658436e-01/(2.092129e-09+1.658436e-01))

```

> EHEX obtained lower nonpareil completeness values than REF and DREX.
> Lowest nonpareil completeness values were ontained for mammals and higest for reptiles.


## Combined community analysis

```{r data_combined, warning=FALSE, comments="", message=FALSE}
species="combined"
genus=species

sample_metadata <- read_tsv(paste0("data/metadata.tsv")) %>%
    rename(dataset=Dataset)

read_counts <- read_tsv("https://sid.erda.dk/share_redirect/BaMZodj9sA/DMB/DMB0134/DMB0134_counts.tsv.gz") %>%
  rename(genome = 1)

genome_metadata <- read_tsv("https://sid.erda.dk/share_redirect/BaMZodj9sA/DMB/DMB0134/DMB0134_mag_info.tsv.gz") %>%
  rename(genome = 1, length=mag_size)

genome_coverage <- read_tsv("https://sid.erda.dk/share_redirect/BaMZodj9sA/DMB/DMB0134/DMB0134_coverage.tsv.gz") %>%
  rename(genome = 1)

download.file("https://sid.erda.dk/share_redirect/BaMZodj9sA/DMB/DMB0134/DMB0134.tree.gz", "data/DMB0134.tree.gz")
genome_tree <- read_tree("data/DMB0134.tree.gz")
```

### Filter data

```{r filter_combined, warning=FALSE, comments="", message=FALSE}
#Filter by coverage
min_coverage=0.3
read_counts_filt <- genome_coverage %>%
  mutate(across(where(is.numeric), ~ ifelse(. > min_coverage, 1, 0))) %>%
  mutate(across(-1, ~ . * read_counts[[cur_column()]]))

# Transform  to genome counts (non-filtered)
readlength=150
genome_counts <- read_counts %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))

# Transform to genome counts (coverage-filtered)
readlength=150
genome_counts_filt <- read_counts_filt %>%
  mutate(across(where(is.numeric), ~ . / (genome_metadata$length / readlength) ))



```

### Alpha diversity

```{r alpha_calotriton, warning=FALSE, comments="", message=FALSE}
#Calculate Hill numbers
richness <- genome_counts_filt %>%
            column_to_rownames(var="genome") %>%
            select(where(~!all(. == 0))) %>%
            hilldiv(.,q=0) %>%
            t() %>%
            as.data.frame() %>%
            rename(richness=1) %>%
            rownames_to_column(var="dataset")

neutral <- genome_counts_filt %>%
            column_to_rownames(var="genome") %>%
            select(where(~!all(. == 0))) %>%
            hilldiv(.,q=1) %>%
            t() %>%
            as.data.frame() %>%
            rename(neutral=1) %>%
            rownames_to_column(var="dataset")

phylogenetic <- genome_counts_filt %>%
            column_to_rownames(var="genome") %>%
            select(where(~!all(. == 0))) %>%
            hilldiv(.,q=1,tree=genome_tree) %>%
            t() %>%
            as.data.frame() %>%
            rename(phylogenetic=1) %>%
            rownames_to_column(var="dataset")

# Merge alpha diversities
alpha_diversity <- richness %>%
      full_join(neutral,by=join_by(dataset==dataset)) %>%
      full_join(phylogenetic,by=join_by(dataset==dataset))

# Write alpha diversities
#alpha_diversity %>% write_tsv(paste0("results/alpha_",genus,".tsv"))

# Print alpha diversity
alpha_diversity %>%
  left_join(sample_metadata %>%
    select(dataset, Extraction, Sample), by = join_by(dataset == dataset)) %>%
  group_by(Extraction) %>%
  summarise(richness = mean(richness), neutral = mean(neutral), phylogenetic = mean(phylogenetic)) %>%
  tt()
```


```{r alpha_diversity_data, warning=FALSE, comments="", message=FALSE}
alpha_data <- alpha_diversity %>%
  left_join(all_data,by= join_by(dataset==dataset))

```

```{r alpha_diversity_plot, warning=FALSE, comments="", message=FALSE, fig.height=6, fig.width=10, fig.fullwidth=TRUE}
alpha_data %>%
    dplyr::select(dataset,Library,Species,Taxon,Sample,Extraction, richness,neutral,phylogenetic) %>% 
    pivot_longer(!c(dataset,Library,Species,Taxon,Sample,Extraction), names_to = "metric", values_to = "value") %>%
    mutate(metric=factor(metric,levels=c("richness","neutral","phylogenetic"))) %>%
    mutate(Taxon=factor(Taxon,levels=c("Amphibian","Reptile","Mammal"))) %>%
    unique() %>%
    ggplot(aes(x=Extraction,y=value, color=Species, group=Extraction)) + 
        geom_boxplot(outlier.shape = NA, fill="#f4f4f4", color="#8c8c8c") + 
        geom_jitter() + 
        scale_color_manual(values=vertebrate_colors) +
        facet_grid(metric ~ Taxon, scales = "free") +
        theme_minimal() +
        labs(y="Diversity",x="Extraction method")
```


### Richness

```{r alpha_diversity_test_richness, warning=FALSE, comments="", message=FALSE}
alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, richness, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(log(richness) ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  plot()

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, richness, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(log(richness) ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  anova()

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, richness, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(log(richness) ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  r.squaredGLMM()

## 2% of the variance explained by fixed effects
## 91% of the variance explained by random effects

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, richness, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(log(richness) ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  VarCorr() %>%
  print(comp = "Variance")

# Within-taxon between-species variance: 59%
91 * (0.85583 / (0.85583 + 0.47587))
# Within-species between-sample variance: 33%
91 * (0.47587 / (0.47587 + 0.85583))
```

> REF got slightly lower richness values than DREX and EHEX.
> Large variability between species within taxa.

### Neutral

```{r alpha_diversity_test_neutral, warning=FALSE, comments="", message=FALSE}
alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, neutral, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(neutral ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  plot()

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, neutral, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(neutral ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  anova()

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, neutral, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(neutral ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  r.squaredGLMM()

## 2% of the variance explained by fixed effects
## 91% of the variance explained by random effects

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, neutral, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(neutral ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  VarCorr() %>%
  print(comp = "Variance")

# Within-taxon between-species variance: 69%
89 * (286.830 / (286.830 + 81.267))
# Within-species between-sample variance: 20%
89 * (81.267 / (286.830 + 81.267))

```

> No significant differences in neutral diversity between extraction methods and taxonomic groups.
> Neutral diversity very species specific.

### Phylogenetic

```{r alpha_diversity_test_phylogenetic, warning=FALSE, comments="", message=FALSE}
alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, phylogenetic, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(phylogenetic ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  plot()

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, phylogenetic, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(phylogenetic ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  anova()

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, phylogenetic, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(phylogenetic ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  r.squaredGLMM()

## 6% of the variance explained by fixed effects
## 90% of the variance explained by random effects

alpha_data %>%
  dplyr::select(dataset, Extraction, Sample, Species, phylogenetic, Taxon, catalogue) %>%
  unique() %>%
  filter(Taxon != "Control") %>%
  lmerTest::lmer(phylogenetic ~ Taxon + Extraction + (1 | Species / Sample), data = ., REML = TRUE) %>%
  VarCorr() %>%
  print(comp = "Variance")

# Within-taxon between-species variance: 43%
90 * (1.37930 / (1.37930 + 1.51968))
# Within-species between-sample variance: 47%
90 * (1.51968 / (1.37930 + 1.51968))
```

> No significant differences in phylogenetic diversity between extraction methods and taxonomic groups.

## Microbial complexity recovery

```{r damr_summary, warning=FALSE, comments="", message=FALSE}
all_data %>%
    select(dataset,Extraction,microbial_fraction,MAG_mapping_percentage,Taxon,catalogue) %>%
    mutate(damr=pmin(1,MAG_mapping_percentage/(microbial_fraction))) %>%
    mutate(damr=ifelse(is.na(damr),0,damr)) %>%
    unique() %>% # what do we filter for?
    group_by(Taxon,Extraction) %>%
    summarise(value = sprintf("%.3f±%.3f", mean(damr), sd(damr))) %>%
    pivot_wider(names_from = Extraction, values_from = value) %>%
    tt(caption = "Mean and standard deviation of breadth of host genome coverage")
```


```{r damr_plot, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
all_data %>%
    select(dataset,Extraction,microbial_fraction,MAG_mapping_percentage,Taxon,Sample,Species) %>%
    mutate(damr=pmin(1,MAG_mapping_percentage/(microbial_fraction))) %>%
    mutate(damr=ifelse(is.na(damr),0,damr)) %>%
    unique() %>%
    ggplot(aes(x=Extraction,y=damr, color=Species, group=Extraction)) + 
        geom_boxplot(outlier.shape = NA, fill="#f4f4f4", color="#8c8c8c") + 
        geom_jitter() + 
        scale_color_manual(values=vertebrate_colors) +
        facet_grid(. ~ Taxon, scales = "free") +
        theme_minimal() +
        labs(y="Domain-adjusted mapping rate",x="Extraction method")
```

```{r damr_test, warning=FALSE, comments="", message=FALSE}
all_data  %>%
    select(dataset,Extraction,microbial_fraction,MAG_mapping_percentage,Taxon, Sample, Species,catalogue) %>%
    mutate(damr=pmin(1,MAG_mapping_percentage/(microbial_fraction))) %>%
    mutate(damr=ifelse(is.na(damr),0,damr)) %>%
    unique() %>%
    filter(Taxon != "Control") %>%
    MASS::glmmPQL(damr~ Taxon+Extraction,random=~1|Species/Sample,
               family="quasibinomial",data=.) %>%
    car::Anova(.,test.statistic="Wald")

data<-all_data  %>%
    select(dataset,Extraction,microbial_fraction,MAG_mapping_percentage,Taxon, Sample, Species,catalogue) %>%
    mutate(damr=pmin(1,MAG_mapping_percentage/(microbial_fraction))) %>%
    mutate(damr=ifelse(is.na(damr),0,damr)) %>%
    unique() %>%
    filter(Taxon != "Control") 


M<-MASS::glmmPQL(damr~ Taxon+Extraction,random=~1|Species/Sample,
               family="quasibinomial",data=data)

Anova_table<-car::Anova(M,test.statistic="Wald",type="III")%>%
    mutate(F=Chisq/Df) # Approximate F value
n <- nrow(M$data)  # Total number of observations
df_fixed <- length(fixef(M))  # Number of fixed effect parameters
df_random <- length(ranef(M))  # Number of random effect parameters (or levels)
approx_residual_df <- n - df_fixed - df_random # Approx. residual Df
data.frame(Anova_table,
           p_value=pf(Anova_table$F, Anova_table$Df, approx_residual_df, lower.tail = FALSE))%>%
  select(F,p_value)

broom.mixed::tidy(M) %>%
    tt()

plot_model(M,type="pred",terms=c("Taxon","Extraction"))

## 27% of the variance explained by fixed effects
## 27% of the variance explained by random effects

VarCorr(M)

# Within-taxon between-species variance: 0%
27*(2.545883e-08/(2.545883e-08+1.911908e+00))
# Within-species between-sample variance: 27%
27*(1.911908e+00/(2.545883e-08+1.911908e+00))
```

> All three extraction methods yield similar DAMRs.


### Community barplot

```{r barplot_combined, warning=FALSE, comments="", message=FALSE, fig.height=8, fig.width=10, fig.fullwidth=TRUE}
# Retrieve EHI taxonomy colors
phylum_colors <- read_tsv("https://raw.githubusercontent.com/earthhologenome/EHI_taxonomy_colour/main/ehi_phylum_colors.tsv") %>%
  right_join(genome_metadata, by=join_by(phylum == phylum)) %>%
    select(phylum, colors) %>%
    unique() %>%
    arrange(phylum) %>%
    select(colors) %>%
    pull()

# Stacked barplot
genome_counts %>%
    mutate_at(vars(-genome), funs(./sum(., na.rm = TRUE)))  %>% #apply TSS nornalisation
    pivot_longer(-genome, names_to = "dataset", values_to = "count") %>%
    left_join(., genome_metadata, by = join_by(genome == genome)) %>% #append taxonomy
    left_join(., sample_metadata, by = join_by(dataset == dataset)) %>%
    mutate(Taxon=factor(Taxon,levels=c("Amphibian",
                                       "Reptile",
                                       "Mammal",
                                       "Bird",
                                       "Control"))) %>%
    mutate(Species=factor(Species,levels=c("Calotriton asper",
                                           "Lissotriton helveticus",
                                           "Salamandra atra",
                                           "Chalcides striatus",
                                           "Natrix astreptophora",
                                           "Podarcis muralis",
                                           "Plecotus auritus",
                                           "Sciurus carolinensis",
                                           "Trichosurus vulpecula",
                                           "Geospizopsis unicolor",
                                           "Perisoreus infaustus",
                                           "Zonotrichia capensis",
                                           "Extraction control",
                                           "Library control"))) %>%
    filter(Taxon != "Control") %>%
    ggplot(., aes(x=dataset,y=count,fill=phylum, group=phylum))+ #grouping enables keeping the same sorting of taxonomic units
          geom_bar(stat="identity", colour="white", linewidth=0.1)+ #plot stacked bars with white borders
          scale_fill_manual(values=phylum_colors) +
          labs(y = "Relative abundance") +
          facet_nested(. ~ Taxon + Species + Sample + Extraction, scales="free_x") +
          guides(fill = guide_legend(ncol = 3)) +
          theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
                          axis.title.x = element_blank(),
                panel.background = element_blank(),
                panel.border = element_blank(),
                panel.grid.major = element_blank(),
                panel.grid.minor = element_blank(),
                axis.line = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
                legend.title=element_blank(),
                panel.spacing = unit(0, "lines"))

#ggsave(paste0("figures/barplot_",genus,".pdf"))
```

```{r filter_dataset, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
genome_counts_filt_variance <- genome_counts_filt %>%
            column_to_rownames(var="genome") %>% 
            select(all_of(sample_metadata %>%
                            filter(!Taxon %in% c("Birds", "Control")) %>% # exclude birds
                            pull(dataset))) %>% 
            select(where(~!all(. == 0)))
```


```{r permanova_combined_richness, warning=FALSE, comments="", message=FALSE}
# Perform betadisper analysis
richness_betadisper_result <- genome_counts_filt_variance %>%
  hillpair(.,q=0, metric="C") %>%
  as.dist() %>% 
  betadisper(., sample_metadata %>% 
                      filter(dataset %in% colnames(genome_counts_filt_variance)) %>%
                      arrange(match(dataset,colnames(genome_counts_filt_variance))) %>% pull(Extraction))
# Check if there are significant differences in dispersion
permutest(richness_betadisper_result,pairwise = TRUE)  # Tests homogeneity of dispersion and pairwise comparisons between groups using permutations.
TukeyHSD(richness_betadisper_result) # test significance between each group


richness <- genome_counts_filt_variance %>%
            hillpair(.,q=0, metric="C") %>%
            as.dist() %>% # nrow(richness) 
            adonis2(formula = . ~ Extraction+Taxon+Species+Sample, 
                    data = sample_metadata %>% 
                      filter(dataset %in% colnames(genome_counts_filt_variance)) %>%
                      arrange(match(dataset,colnames(genome_counts_filt_variance))), 
                    permutations = 999, 
                    by="terms") %>%
            broom::tidy()
richness
```

```{r permanova_combined_neutral, warning=FALSE, comments="", message=FALSE}
# Perform betadisper analysis
neutral_betadisper_result <- genome_counts_filt_variance %>%
  hillpair(.,q=1, metric="C") %>%
  as.dist() %>% 
  betadisper(., sample_metadata %>% 
                      filter(dataset %in% colnames(genome_counts_filt_variance)) %>%
                      arrange(match(dataset,colnames(genome_counts_filt_variance))) %>% pull(Extraction))
# Check if there are significant differences in dispersion
permutest(neutral_betadisper_result,pairwise=TRUE)  # Tests homogeneity of dispersion and pairwise comparisons between groups using permutations.

neutral <- genome_counts_filt_variance %>%
            hillpair(.,q=1, metric="C") %>%
            as.dist() %>%
            adonis2(formula = . ~ Extraction+Taxon+Species+Sample, 
                    data = sample_metadata %>% 
                      filter(dataset %in% colnames(genome_counts_filt_variance)) %>%
                      arrange(match(dataset,colnames(genome_counts_filt_variance))), 
                    permutations = 999, 
                    by="terms") %>%
            broom::tidy()
neutral
```


```{r permanova_combined_phylogenetic, warning=FALSE, comments="", message=FALSE}
# Perform betadisper analysis
phylogenetic_betadisper_result <- genome_counts_filt_variance %>%
  hillpair(.,q=1, metric="C") %>%
  as.dist() %>% 
  betadisper(., sample_metadata %>% 
                      filter(dataset %in% colnames(genome_counts_filt_variance)) %>%
                      arrange(match(dataset,colnames(genome_counts_filt_variance))) %>% pull(Extraction))
# Check if there are significant differences in dispersion
permutest(phylogenetic_betadisper_result,pairwise = TRUE)  # Tests homogeneity of dispersion and pairwise comparisons between groups using permutations.

phylogenetic <- genome_counts_filt_variance %>%
            hillpair(.,q=1, tree=genome_tree, metric="C") %>%
            as.dist() %>%
            adonis2(formula = . ~ Extraction+Taxon+Species+Sample, 
                    data = sample_metadata %>% 
                      filter(dataset %in% colnames(genome_counts_filt_variance)) %>%
                      arrange(match(dataset,colnames(genome_counts_filt_variance))), 
                    permutations = 999, 
                    by="terms") %>%
            broom::tidy()
phylogenetic

```

```{r variance_partitioning_barplot, warning=FALSE, comments="", message=FALSE, fig.height=6, fig.width=4}
data.frame(
  Term = factor(rep(c("Extraction", "Taxon", "Species","Sample","Residual"), times = 3),levels = c("Extraction", "Taxon", "Species","Sample","Residual")),
  Metric = factor(rep(c("Richness", "Neutral", "Phylogenetic"), each = 5),levels = c("Richness", "Neutral", "Phylogenetic")),
  Value = c(richness$R2[-6],neutral$R2[-6],phylogenetic$R2[-6])
)%>%
  ggplot(., aes(x = Metric, y = Value, fill = Term)) +
  geom_bar(stat = "identity", position = "stack") +
  #scale_fill_brewer(palette = "Set1") +  # Choose a color palette
  scale_fill_manual(values = c("#47bfb6", "#4a7015", "#5e1717", "#9e6b24", "#4a4a4a")) +
  labs(
    x = "Beta diversity metric",
    y = "R-square",
    fill = "Term"
  ) +
  theme_minimal()

```

```{r nmds_combined, warning=FALSE, comments="", message=FALSE, fig.height=4, fig.width=10, fig.fullwidth=TRUE}
genome_counts_NMDS <- genome_counts_filt_variance %>%
            #column_to_rownames(var="genome") %>%
            select(where(~!all(. == 0))) %>%
            hillpair(.,q=1, metric="C", out="dist") %>%
            metaMDS(.,trymax = 999, k=2, trace=0) %>%
            vegan::scores() %>%
            as_tibble(., rownames = "dataset") %>%
            left_join(sample_metadata, by = join_by(dataset == dataset)) %>%
            filter(Taxon != "Control") %>%
            group_by(Sample) %>%
            mutate(sample_x=mean(NMDS1), sample_y=mean(NMDS2))

genome_counts_NMDS %>%
            ggplot(., aes(x=NMDS1,y=NMDS2, color=Species, shape=Extraction)) +
                scale_color_manual(values=vertebrate_colors) +
                geom_point(size=3, alpha=0.8) +
                geom_segment(aes(x=sample_x, y=sample_y, xend=NMDS1, yend=NMDS2), alpha=0.2) +
                theme_classic() +
                theme(legend.position="right", legend.box="vertical") +
                guides(color=guide_legend(title="Species"))
```
